A lottery model of density-dependent selection in evolutionary genetics

### 2017-05-04 manuscript.tex
latest

Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate the direction of evolution for each. To proceed, we map these verbally-defined environments to quantitative parameter regimes in our model, summarized in Fig.~\ref{fig:table}.

@@

Disturbed environments are characterized by short bursts of high extrinsic mortality $d$ caused by physical destruction. We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ (rather than $\Delta_+ n_i$) territories are secured by type $i$ each iteration. In the heavily disturbed extreme of Grime's triangle, many such bursts (which could each be spatially localized) will occur each iteration (i.e. over the time required for a propagule to reach reproductive maturity), which we approximate as a high constant mortality rate $d\approx 1$. Population persistence requires $b>d/(1-d)\gg 1$ (Appendix C), i.e. severe disturbance comes with a high fecundity/dispersal requirement. We expect that it is not possible to greatly exceed this requirement ($b\approx d/(1-d)$), which implies $L\ll 1$.

Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein). Severe stress inhibits growth and reproduction, so that $b\ll 1$. In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$; coping with the environmental stress is barely possible \citep{grime_1974,grime_1977}. In our model, this implies $L\ll 1$.

@@

The expected direction of evolution is determined by three factors. First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them. Second, adaptive mutations in one trait may occur more frequently than in others \citep{yampolsky_2001}. We do no doubt that these factors operate in nature and contribute towards ``primary strategy'' trait patterns. However, modeling how mutations affect traits subject to trade-offs is beyond the scope of this article, and so we will not evaluate these first two factors: we assume that mutations affect $b$, $c$ and $d$ independently, and that $b$, $c$ and $d$ have the same distribution of beneficial fitness effects.

The third factor is that some traits may have a greater impact on fitness: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}. Very few neutral mutations fix (probability of order $1/N$).

@@

To represent Grime's disturbed environment, $\Delta_+ n_i$ is replaced with $(1-d_i)\Delta_+ n_i$, and a single type reaches equilibrium when $L/(1-e^{-L})=(1-d_i)b_i/d_i$ (population persistence now requires $(1-d_i)b_i>d_i$ rather than $b_i>d_i$). We assumed $d_i\approx 1$ and $(1-d_i)b_i\approx d_i$, so that $L\ll 1$. A mutant's invasion fitness is then given by
\begin{equation}
\Delta n_j/n_j \approx (1-d_j)b_j - d_j.
\end{equation}
Another distinct feature of disturbed environments is that the bulk of mortality is environmental and unavoidable. That is, $d_i=d_{ie}+d_{ig}$, where the environmental part $d_{ie}\approx 1$ does not evolve, and the genetic part $d_{ig}\ll 1$ is affected by mutations in the same way as before $d_{jg}=(1-\epsilon) d_{ig}$.

The effects of improving $b$ and $d$ are now no longer identical for given values of $d_i$ and $d_{ig}$. Increasing $b$ by a fraction $\epsilon$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon d_i$, whereas reducing $d_g$ by the same fraction gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$.

Not surprisingly, selection for lower $d$ is weaker than for greater $b$ if $d_{ig}$ is sufficiently small. However, assuming that there is some non-negligible genetic contribution to mortality

Thus, since $d_i\approx 1$, selection for lower $d$ is potentially much stronger than that for higher $b$.

even though the genetic component of mortality $d_{g}$ is small compared to the environmental component $d_{e}$, if disturbance is intense enough ( is small enough), .

>>>>

Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate role of selection in shaping the corresponding ``primary strategies''. To proceed, we map these verbally-defined environments to quantitative parameter regimes in our model, summarized in Fig.~\ref{fig:table}.

@@

Disturbed environments are characterized by short bursts of high extrinsic mortality $d$ caused by physical destruction. We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ (rather than $\Delta_+ n_i$) territories are secured by type $i$ each iteration. In the heavily disturbed extreme of Grime's triangle, many such bursts (which could each be spatially localized) will occur each iteration (i.e. over the time required for a propagule to reach reproductive maturity), which we approximate as a high constant mortality rate $d\approx 1$. Population persistence requires $b>d/(1-d)$, which implies $b\gg 1$ (severe disturbance comes with a high fecundity/dispersal requirement). We assume this requirement can just be met, so that $b\approx d/(1-d)$ and $L\ll 1$ (Appendix C).

Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein). Severe stress inhibits growth and reproduction, so that $b\ll 1$. In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$; it is just possible to cope with the environmental stress \citep{grime_1974,grime_1977}. In our model, this implies $L\ll 1$.

@@

The expected direction of evolution is determined by three factors. First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them. Second, adaptive mutations in one trait may occur more frequently than in others \citep{yampolsky_2001}. Third, selection favors some trait changes over others: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}. Only a tiny proportion of neutral and deleterious mutations fix (probability of order $1/N$ for neutrals).

Here we only consider the last of these factors. To account for the first two factors would require an evaluation of the effects of mutations on traits subject to trade-offs, which is beyond the scope of this article. Thus, we consider how selection alone shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with the same distribution of beneficial fitness effects.

@@

To represent Grime's disturbed environment, $\Delta_+ n_i$ is replaced with $(1-d_i)\Delta_+ n_i$, and a single type reaches equilibrium when $L/(1-e^{-L})=(1-d_i)b_i/d_i$ (population persistence now requires $(1-d_i)b_i>d_i$ rather than $b_i>d_i$). We assumed $d_i\approx 1$ and $(1-d_i)b_i\approx d_i$, so that $L\ll 1$. A mutant's invasion fitness is then given by
\begin{equation}
\Delta n_j/n_j \approx (1-d_j)b_j - d_j.
\end{equation}
Another distinct feature of disturbed environments is that the bulk of mortality is environmental and unavoidable. That is, $d_i=d_{ie}+d_{ig}$, where the environmental part $d_{ie}$ does not evolve, and the genetic part $d_{ig}$ is affected proportionally by mutations $d_{jg}=(1-\epsilon) d_{ig}$.

The effects of improving $b$ and $d$ are now no longer identical for given values of $d_{ie}$ and $d_{ig}$. A $b-$mutation $b_j=(1+\epsilon)b_i$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon d_i \approx \epsilon$, whereas a $d$-mutation $d_{jg}=(1-\epsilon) d_{ig}$ gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$. Since overall mortality is high $d_i\approx 1$, $d_{g}$ can be small compared to $d_{e}$ and yet still make a greater contribution to invasion fitness than $b$. The case where $d_{ig}$ is so small that $d_{ig}\ll 1-d_i$ (i.e. $d_{ig}$ is an order of magnitude smaller than $1-d_i$ which itself an order of magnitude smaller than $d_{ie}$) effectively represents a hard constraint that $d$ cannot evolve. We do not deny this possibility, but this bring us back to the issue of mutations affect traits (the first two factors controlling the direction of evolution in section ``Primary strategies and Grime's triangle''), which we do not address in this manuscript.

Thus, assuming that $d$ can evolve in this sense, $d_{ig}$ is either comparable to or greater than $1-d_i$. Yet $d_{ig}$ cannot be appreciably greater than $1-d_i$ for long, because selection will then strongly favor reductions in $d_g$ over increases in $b$, thereby bringing $d_{ig}$ closer to $1-d_i$. Thus, we expect selection on $b$ and $d$ to be of comparable strength in disturbed environments as well.
